---
title: "Georgia Recidivism Prediction"
author: "Junyi Yang, Ziyi Guo, Jiewen Hu"
date: "Apirl, 2024"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```


```{r load_packages, warning = FALSE}
options(scipen=10000000)

library(tidyverse)
library(kableExtra)
library(caret)
library(knitr) 
library(pscl)
library(plotROC)
library(pROC)
library(lubridate)
library(RSocrata)
```

```{r load_data, cache = TRUE}
palette5 <- c("#981FAC","#CB0F8B","#FF006A","#FE4C35","#FE9900")
palette4 <- c("#981FAC","#FF006A","#FE4C35","#FE9900")
palette2 <- c("#981FAC","#FF006A")


# Read and process burglaries data
Recidivism <- 
  read.socrata("https://data.ojp.usdoj.gov/Courts/NIJ-s-Recidivism-Challenge-Full-Dataset/ynf5-u8nk/") %>% 
  na.omit()  # Remove rows with missing values


```

# Data Exploration

## Discussion:

**The trade-off between sensitivity and specificity**

Sensitivity (also known as the true positive rate or recall) refers to the model's ability to correctly identify repeat offenders. Specificity (also known as the true negative rate) refers to the model's ability to correctly identify non-repeat offenders. In the criminal justice system, prioritizing sensitivity means we are more likely to identify potential repeat offenders, but this can also lead to more false positives (i.e., incorrectly labeling someone as likely to reoffend). Prioritizing specificity, on the other hand, reduces the number of false positives but increases the risk of false negatives (i.e., failing to identify someone who will actually reoffend).

**The costs and consequences of prioritizing sensitivity**

This could lead to more people being unfairly monitored or retained in prison, not only affecting individual freedom but also increasing social and economic costs, especially for lower socioeconomic and marginalized groups.

**The costs and consequences of prioritizing specificity**

This could lead to more individuals with a risk of reoffending being released, increasing the risk of societal recidivism, but it also reduces unfair treatment of individuals and social costs.

```{r add_feature}
# filter the crime data under 3 years

Recidivism <- Recidivism %>%
  mutate(Recidivism_numeric = ifelse(recidivism_within_3years == "true", 1, 0))
glimpse(Recidivism)
```




```{r exploratory_continuous, fig.width=10}

Recidivism %>%
  dplyr::select(recidivism_within_3years, jobs_per_year, percent_days_employed, supervision_risk_score_first, avg_days_per_drugtest, drugtests_thc_positive, drugtests_cocaine_positive, drugtests_meth_positive, drugtests_other_positive) %>%
  gather(Variable, value, -recidivism_within_3years) %>%
    ggplot(aes(recidivism_within_3years, value, fill=recidivism_within_3years)) + 
      geom_bar(position = "dodge", stat = "summary", fun = "mean") + 
      facet_wrap(~Variable, scales = "free", ncol = 4) +
      scale_fill_manual(values = palette2) +
      labs(x="recidivism_within_3years", y="Value", 
           title = "Feature associations with the likelihood of recidivism within 3 years",
           subtitle = "(continous outcomes)") +
      theme_minimal() + theme(legend.position = "none")
```

```{r exploratory_continuous_density, fig.width=10, message=FALSE, warning=FALSE}

Recidivism %>%
    dplyr::select(recidivism_within_3years, jobs_per_year, percent_days_employed, supervision_risk_score_first, avg_days_per_drugtest, drugtests_thc_positive, drugtests_cocaine_positive, drugtests_meth_positive, drugtests_other_positive) %>%
    gather(Variable, value, -recidivism_within_3years) %>%
    ggplot() + 
    geom_density(aes(value, color=recidivism_within_3years), fill = "transparent") + 
    facet_wrap(~Variable, scales = "free", ncol = 4) +
    scale_colour_manual(values = palette2) +
    labs(x="Value", y="Density",
         title = "Feature Distribution with the likelihood of recidivism within 3 years",
         subtitle = "(continous outcomes)") +
      theme_minimal() + theme(legend.position = "none")

```


```{r exploratory_binary, fig.width=10, message=FALSE, warning=FALSE}
Recidivism %>%
    dplyr::select(recidivism_within_3years, gender,race, age_at_release, education_level, education_level, dependents, prison_offense) %>%
    gather(Variable, value, -recidivism_within_3years) %>%
    count(Variable, value, recidivism_within_3years) %>%
      ggplot(., aes(value, n, fill = recidivism_within_3years)) +   
        geom_bar(position = "dodge", stat="identity") +
        facet_wrap(~Variable, scales="free") +
        scale_fill_manual(values = palette2) +
        labs(x="recidivism_within_3years", y="Value",
             title = "Feature associations with the likelihood of recidivism within 3 years",
             subtitle = "Categorical features") +
        theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
      theme_minimal() + theme(legend.position = "none")
```

# Create A Logistic Regression Model


```{r create_partition}
set.seed(3456)
trainIndex <- createDataPartition(Recidivism$recidivism_within_3years, p = .50,
                                  list = FALSE,
                                  times = 1)
RecidivismTrain <- Recidivism[ trainIndex,]
RecidivismTest  <- Recidivism[-trainIndex,]

```


```{r run_model}

RecidivismModel <- glm(Recidivism_numeric ~ .,
                  data=RecidivismTrain %>% 
                    dplyr::select(Recidivism_numeric,jobs_per_year,percent_days_employed, supervision_risk_score_first, prison_years, condition_cog_ed,condition_mh_sa, condition_other, gang_affiliated,prior_arrest_episodes_violent, prior_arrest_episodes_property, violations, violations_instruction, violations_failtoreport, violations_1, delinquency_reports,program_attendances, program_unexcusedabsences, residence_changes,  prior_arrest_episodes_drug,prior_arrest_episodes,prior_revocations_parole,prior_revocations_probation, avg_days_per_drugtest,drugtests_thc_positive,drugtests_cocaine_positive,drugtests_meth_positive,drugtests_other_positive,gender,race, age_at_release, residence_puma,education_level, dependents, prison_offense),
                  family="binomial" (link="logit"))

Recidivism_sum <- summary(RecidivismModel)

coefficients_table <- as.data.frame(Recidivism_sum$coefficients)

coefficients_table$significance <- ifelse(coefficients_table$`Pr(>|z|)` < 0.001, '***',
                                         ifelse(coefficients_table$`Pr(>|z|)` < 0.01, '**',
                                                ifelse(coefficients_table$`Pr(>|z|)` < 0.05, '*',
                                                       ifelse(coefficients_table$`Pr(>|z|)` < 0.1, '.', ''))))

coefficients_table$p_value <- paste0(round(coefficients_table$`Pr(>|z|)`, digits = 3), coefficients_table$significance)

coefficients_table %>%
  select(-significance, -`Pr(>|z|)`) %>% 
  kable(align = "r") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))  %>%
  footnote(general_title = "\n", general = "Table x")
```


1.*gang affiliated*
Gang affiliation significantly increases the risk of reoffending.

2.*prison years*
Serving a prison sentence of more than 2 to 3 years is associated with a reduced risk of recidivism, while sentences of less than one year may lead to a higher risk.

3.*employment*
individuals who have more jobs in a year are more likely to engage in recidivism, possibly due to instability or low job quality.individuals who are employed for a higher percentage of days are less likely to engage in recidivism, implying that stable employment may be a protective factor against recidivism.

4.*programs*
high attendance in rehabilitation or support programs may be positively correlated with lower recidivism rates, as it could indicate an individual's commitment to rehabilitation. Unexcused absences might be negatively correlated with successful rehabilitation, potentially indicating a lack of engagement with the program, which could be a risk factor for re-offending.

5.*drug*
Positive drug tests, especially for THC and methamphetamine, are associated with an increased likelihood of recidivism.It could reflect the severity of drug habits or other social and economic factors tied to drug use that influence the likelihood of re-engagement in criminal behavior. 



```{r fit_metrics}

pR2(RecidivismModel)

```
McFadden's pseudo-R-squared of approximately 0.1897, or 18.97%, indicates that the model explains a moderate proportion of the variance in recidivism. This value suggests that while the model has some explanatory power, there are other factors not included in the model that also affect the likelihood of recidivism.

# Make Predictions

We create a dataframe of predictions for the 500 observations in our test set, called `testProbs`.

These predictions are the estimated probabilities of clicking for these out-of-sample subjects. We can compare them to the observed outcome.

Run the code below and explore using `glimpse(testProbs)` to see what these predictions look like.

```{r testProbs}

testProbs <- data.frame(Outcome = as.factor(RecidivismTest$Recidivism_numeric),
                        Probs = predict(RecidivismModel, RecidivismTest, type= "response"),
                        gender = RecidivismTest$gender,
                        race = RecidivismTest$race)

head(testProbs)
```

## Discussion 3

Look at the plot of our predicted probabilities for our observed clickers (`1`) and non-clickers (`0`). **Write a sentence or two about how you think our model is performing.**

```{r plot_testProbs}
ggplot(testProbs, aes(x = Probs, fill = as.factor(Outcome))) + 
  geom_density() +
  facet_grid(Outcome ~ .) +
  scale_fill_manual(values = palette2) +
  labs(x = "Click", y = "Density of probabilities",
       title = "Distribution of predicted probabilities by observed outcome") +
  theme(strip.text.x = element_text(size = 18),
        legend.position = "none") +
      theme_minimal() + theme(legend.position = "none")
```

this chart shows a promising degree of discriminating power in the model, as evidenced by the presence of peaks suggesting a concentration of true positives (red area towards the right) and true negatives (purple area towards the left). Although there is some overlap between the predicted probabilities for the two classes, which suggests areas where the model's predictions are less certain, the separation between the peaks indicates the model is capturing a meaningful difference between the outcomes. 


# Confusion Matrix

Each threshold (e.g. a probability above which a prediction is a "click" and below which it's a "no click") has it's own rate of error. These errors can be classified in four ways for a binary model.

A "confusion matrix" for the threshold of 50% shows us the rate at which we got True Positives (aka Sensitivity), False Positives, True Negatives (aka Specificity) and False Negatives for that threshold.

```{r }
testProbs <- 
  testProbs %>%
  mutate(predOutcome  = as.factor(ifelse(testProbs$Probs > 0.5 , 1, 0)),
         error = ifelse(testProbs$predOutcome == testProbs$Outcome, 0, 1)) 
race_difference <- testProbs %>% 
  group_by(race, gender) %>%
  summarize(total_error = sum(error),
            total_people = n()) %>%
  mutate(percent_error = total_error / total_people) 
race_difference %>% 
  kable(align = "r") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))  %>%
  footnote(general_title = "\n", general = "Table x")
```

```{r confusion_matrix}
cm <- caret::confusionMatrix(testProbs$predOutcome, testProbs$Outcome, 
                       positive = "1")

print(cm)


mosaicplot(cm$table, color=c("#FF006A","#FE9900"), main = "Mosaic Plot for Original Confusion Matrix",
           xlab = "Prediction", ylab = "Reference")


```

The model correctly predicts the outcome 71.03% of the time and We can be 95% confident that the true accuracy of the model is between 70.12% and 71.93%. p value and kappa value also suggest that accuracy of the model is better than what would be achieved by always predicting the most frequent class.

Sensitivity measures the proportion of actual positive cases (recidivism) that the model correctly identifies. A sensitivity of 0.7845, or 78.45%, means that out of all the individuals who did recidivist, the model correctly identified approximately 78.45% of them. This is a fairly high rate, indicating the model's strength in capturing those at risk of recidivism. Specificity measures the proportion of actual negative cases (non-recidivism) that the model correctly identifies. A specificity of 0.6094, or 60.94%, indicates that out of all the individuals who did not recidivist, the model correctly identified about 60.94% of them as not being at risk. The specificity is lower than the sensitivity, which suggests that the model is somewhat less adept at correctly identifying those who will not recidivist.

In conclusion, the model is effective at identifying individuals who will recidivate but could benefit from improvements in accurately identifying those who will not. 

## Discussion 4

**Describe what each of the following mean in the context of our advertising use case:**

**True Positive:** We predicted a click and it was a click IRL

**False Positive:**

**True Negative:**

**False Negative:**

# ROC Curve

The ROC curve, gives us another visual "goodness of fit" metric. One that is a bit more tricky. You want to have a curve that is "above" the y=x line, which is where your prediction rates for positives and negatives are "no better than a coin flip". If it's too "square" - you are probably over fit. The Area-Under-The-Curve or "AUC" calculation below will help guide your understanding of the ROC curve

```{r auc, message = FALSE, warning = FALSE}
auc(testProbs$Outcome, testProbs$Probs)
```
An AUC of 0.7747 means that there is a 77.47% chance that the model will be able to distinguish between a randomly chosen positive instance (one that actually did recidivate) and a negative instance (one that did not recidivate).

```{r roc_curve, warning = FALSE, message = FALSE}
ggplot(testProbs, aes(d = as.numeric(Outcome), m = Probs)) +
  geom_roc(n.cuts = 50, labels = FALSE, colour = "#FF006A") +
  style_roc(theme = theme_grey) +
  geom_abline(slope = 1, intercept = 0, size = 1.5, color = 'grey') +
  labs(title = "ROC Curve - clickModel") +
      theme_minimal() + theme(legend.position = "none")
```

The curve rises steeply towards the upper-left corner of the plot, which shows that the model has a strong true positive rate before accruing false positives.

## Discussion 5

Try to come up with an explanation of what this ROC curve is "saying" in 1-2 sentences.

Is it useful? Is it overfit? What does the y=x line represent?

# Cross validation

We run 100-fold cross validation and look at the ROC (aka AUC), Sensitivity and Specificity across this series of predicitons. How do they look?

Probably pretty, pretty good.

```{r cv}
ctrl <- trainControl(method = "cv", number = 100, classProbs=TRUE, summaryFunction=twoClassSummary)

cvFit <- train(recidivism_within_3years ~ .,
                  data=Recidivism %>% 
                  dplyr::select(recidivism_within_3years,jobs_per_year,percent_days_employed, supervision_risk_score_first, prison_years, condition_cog_ed,condition_mh_sa, condition_other, gang_affiliated,prior_arrest_episodes_violent, prior_arrest_episodes_property, violations, violations_instruction, violations_failtoreport, violations_1, delinquency_reports,program_attendances, program_unexcusedabsences, residence_changes,  prior_arrest_episodes_drug,prior_arrest_episodes,prior_revocations_parole,prior_revocations_probation, avg_days_per_drugtest,drugtests_thc_positive,drugtests_cocaine_positive,drugtests_meth_positive,drugtests_other_positive,gender,race, age_at_release, residence_puma,education_level, dependents, prison_offense), 
                method="glm", family="binomial",
                metric="ROC", trControl = ctrl)

cvFit
```

The Area Under the Curve (AUC) from the cross-validated model is 0.7751, which is slightly higher than the previously mentioned AUC of 0.7747. This suggests that the model's ability to discriminate between the positive and negative classes is consistent and robust across different subsets of the data.

The cross-validated sensitivity is lower than the previously mentioned sensitivity of 0.7845. This indicates that, across various folds, the model's ability to correctly identify true positives (actual cases of recidivism) is somewhat less than what was observed in the initial result. The specificity is higher than the initially mentioned specificity of 0.6094. This implies that the model has a better ability to correctly identify true negatives (actual cases of non-recidivism) when evaluated under the cross-validation process.

Comparatively, the cross-validated results provide a more reliable estimate of the model's performance since cross-validation reduces bias from any potential overfitting to a single test set. The differences in the cross-validated sensitivity and specificity compared to the initial results suggest that the model may be more conservative in predicting positive cases (recidivism) but is quite robust in correctly identifying negative cases (non-recidivism) across different subsamples of the dataset.

```{r goodness_metrics, message = FALSE, warning = FALSE}
dplyr::select(cvFit$resample, -Resample) %>%
  gather(metric, value) %>%
  left_join(gather(cvFit$results[2:4], metric, mean)) %>%
  ggplot(aes(value)) + 
    geom_histogram(bins=35, fill = "#FF006A", alpha=0.6) +
    facet_wrap(~metric) +
    geom_vline(aes(xintercept = mean), colour = "#981FAC", linetype = 2, size = 0.9) +
    scale_x_continuous(limits = c(0, 1)) +
    labs(x="Goodness of Fit", y="Count", title="CV Goodness of Fit Metrics",
         subtitle = "Across-fold mean reprented as dotted lines") +
      theme_minimal() + theme(legend.position = "none")

```

# Cost-Benefit Calculation

This has all been building to an examination of the model in the context of our ad campaign. Let's set this up to estimate the revenues associated with using this model under the following scenario:

-An impression (serving an ad) costs \$0.10

-A click brings an estimated \$0.35 of revenue per visitor on average.

## Discussion 6

Run the code below and look at the revenues associated with each prediction type. (Notice our `Revenue` calculation - for your assignment this will more closely resemble the calculation in the text book).

**What is the rate of return per dollar spent?**

A clue to figuring this out - we only spend money for impressions on True Positives and False Positives, and we lost money with our False Negatives - hypothetically we spend nothing on True Negatives.

**Are there particular types of error which are more damaging? What outcomes do we want to maximize or minimize? Why do we look at False Negatives as a negative cost?**



States spent an average of $45771 per prisoner for the year.
https://usafacts.org/articles/how-much-do-states-spend-on-prisons/

https://csgjusticecenter.org/publications/the-cost-of-recidivism/
8000000000/193000 = 41450.78
```{r cost_benefit}
cost_benefit_table <-
   testProbs %>%
      count(predOutcome, Outcome) %>%
      summarize(True_Negative = sum(n[predOutcome==0 & Outcome==0]),
                True_Positive = sum(n[predOutcome==1 & Outcome==1]),
                False_Negative = sum(n[predOutcome==0 & Outcome==1]),
                False_Positive = sum(n[predOutcome==1 & Outcome==0])) %>%
       gather(Variable, Count) %>%
       mutate(Revenue =
               ifelse(Variable == "True_Negative", Count * 0,
               ifelse(Variable == "True_Positive",((45771) * Count),
               ifelse(Variable == "False_Negative", (41450.78) * Count,
               ifelse(Variable == "False_Positive", (-45771) * Count, 0))))) %>%
    bind_cols(data.frame(Description = c(
              "We correctly predicted no recidivism",
              "We correctly predicted recidivism",
              "We predicted no recidivism but get recidivism",
              "We predicted recidivism but get no recidivism")))

kable(cost_benefit_table,
       caption = "Cost/Benefit Table") %>% kable_styling()
```

# Optimize Thresholds

The last step to tuning our model is to run it for each threshold value. Recall that we chose 0.5 as the line above which a prediction is classified as a "click". We can then look at the confusion matrices for each threshold and choose the one that returns the most revenue.

The code below bakes in our cost-revenue calculations.

## Discussion 7

**Consider how revenues compare to a situation in which we had no model (e.g. served adds to all 1000 in the sample) or one in which we don't tune the threshold.**

```{r iterate_threshold}
iterateThresholds <- function(data) {
  x = .01
  all_prediction <- data.frame()
  while (x <= 1) {
  
  this_prediction <-
      testProbs %>%
      mutate(predOutcome = ifelse(Probs > x, 1, 0)) %>%
      count(predOutcome, Outcome) %>%
      summarize(True_Negative = sum(n[predOutcome==0 & Outcome==0]),
                True_Positive = sum(n[predOutcome==1 & Outcome==1]),
                False_Negative = sum(n[predOutcome==0 & Outcome==1]),
                False_Positive = sum(n[predOutcome==1 & Outcome==0])) %>%
     gather(Variable, Count) %>%
     mutate(Revenue =
               ifelse(Variable == "True_Negative", Count * 0,
               ifelse(Variable == "True_Positive",((.35 - .1) * Count),
               ifelse(Variable == "False_Negative", (-0.35) * Count,
               ifelse(Variable == "False_Positive", (-0.1) * Count, 0)))),
            Threshold = x)
  
  all_prediction <- rbind(all_prediction, this_prediction)
  x <- x + .01
  }
return(all_prediction)
}
```

```{r revenue_model}
whichThreshold <- iterateThresholds(testProbs2)

whichThreshold_revenue <- 
whichThreshold %>% 
    group_by(Threshold) %>% 
    summarize(Revenue = sum(Revenue))

  ggplot(whichThreshold_revenue)+
  geom_line(aes(x = Threshold, y = Revenue))+
  geom_vline(xintercept =  pull(arrange(whichThreshold_revenue, -Revenue)[1,1]))+
    labs(title = "Model Revenues By Threshold For Test Sample",
         subtitle = "Vertical Line Denotes Optimal Threshold") +
      theme_minimal()

```



